Modeling mental rotation

### 2014-01-28 man/cogsci-2014/cogsci-2014.tex
Some work on background

\title{An active sampling model of mental rotation}

@@

@@

One of our most astonishing cognitive feats is the ability to
envision, manipulate, and plan with objects -- all without actually
perceiving them. This capacity for ``mental simulation'' has been
widely studied and has sparked intense debate about the underlying
representation of mental images \cite{Kosslyn:2009tj,
  Pylyshyn:2002vk}. Despite this vast literature, there have been
surprisingly few computational models of mental imagery.

@@

classic result by shepard and metzler

previous models of mental imagery: \cite{Kosslyn:1977tv, Anderson1978,
  Funt:1983wn, Glasgow:1992tj}

assume people rotate until congruent

but how do people actually construct the plan for the rotation?

evidence that people *don't* actually rotate until congruent,
e.g. rotating away from near-perfect matches \cite{Gardony:2013gn}

@@

@@

@@

Formally, we denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is
generated by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$,
where $\theta$ is a rotation, $h=0$ is the hypothesis that the images
depict the same object, and $h=1$ is the hypothesis that the images
depict mirror-image objects. The posterior probability of each
hypothesis given the observed shapes is then:
\begin{equation}
  p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta, h)p(X_a)p(h)p(\theta)\dif\theta.
  \label{eq:posterior}
\end{equation}
Because we ultimately want to determine which hypothesis is more
likely, the quantity of interest is a ratio:
\begin{equation*}
  \ell := \frac{\int p(X_b\ \vert\ X_a, \theta, \hi)p(X_a)p(\hi)p(\theta)\dif\theta}{\int p(X_b\ \vert\ X_a, \theta, \hf)p(X_a)p(\hf)p(\theta)\dif\theta},
\end{equation*}
which (assuming the two hypotheses are equally likely \textit{a
  priori}, and that all rotations are equally likely) simplifies to:
\begin{equation}
  \ell = \frac{\int p(X_b\ \vert\ X_a, \theta, \hi)\dif\theta}{\int p(X_b\ \vert\ X_a, \theta, \hf)\dif\theta}.
  \label{eq:lh-ratio}
\end{equation}
If $\ell > 1$, then we accept the hypothesis that the images depict
the same object ($\hi$); if $\ell < 1$, then we accept the hypothesis
that the images depict flipped objects ($\hf$).

@@

The previous hypothesis in the literature is that people perform a
mental rotation in the shortest direction until a match is found, and
that determining this direction is an implementation detail \TODO{cite
  anderson?}. To reflect this hypothesis, we created an ``oracle''
model which knows the correct rotation \textit{a priori}, and simply
performs rotations in the shortest direction until it reaches the
correct rotation.

@@

Unfortunately, it is not always clear what constitutes a good enough
``match''. Many shapes may have multimodal similarity functions
\TODO{refer to figure}, and the maximum possible value of $S$ is
unlikely to be known \textit{a priori} \TODO{cite something about how
  shape matching is hard}. In this case, we must weaken our stopping
criteria: rather than searching for a value above a threshold, we
search for \textit{any} local maximum, as in the Hill Climbing search
algorithm (\Hc{}). Similar to the \Th{} model, we use the (local) MLE
estimate of $\theta$, as in Eq.~\ref{eq:mle-lh-ratio}.

@@

An alternative to the previous models is not to perform a search for
the MLE, but to approximate the integrals in Eq.~\ref{eq:lh-ratio} by
picking locations to sample which will most improve accuracy of the
estimate. Rather than a point estimate of the ratio, or even a simple
Monte Carlo estimate, we can infer a distribution by placing priors
over the similarity functions themselves:
\begin{equation}
\ell = \frac{\int \left[\int S(\Xb, f(\Xa, \theta, \hi))p(\theta)\dif\theta\right] p(S)\dif S}{\int \left[\int S(\Xb, f(\Xa, \theta, \hf))p(\theta)\dif\theta\right] p(S)\dif S}
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use Gaussian Process (GP) priors, as in \citeA{Osborne:2012tm}
\TODO{probably need to describe better what a GP is}. Because our
domain is on a circle, we utilize a periodic kernel for the covariance
\cite{Rasmussen:2006vz}, i.e. $k(\theta,
\theta^\prime)=h^2\exp[-\frac{2}{w^2}\sin^2(\frac{1}{2}(\theta-\theta^\prime))]$.

@@

Revisiting our likelihood ratio, we now have:
\begin{equation}
p(\ell)\approx\frac{\mathcal{N}(\mathbb{E}[Z_0], \mathbb{V}(Z_0))}{\mathcal{N}(\mathbb{E}[Z_1], \mathbb{V}(Z_1))}.
\end{equation}
This distribution does not have a nice form. However, we really are
just interested in whether $Z_0>Z_1$ or $Z_1>Z_0$: thus, rather than
computing $p(\ell)$, we use $Z_D=Z_0-Z_1$. Assuming $Z_0$ and $Z_1$
are independent,\footnote{This assumption is probably not
  correct. \TODO{explanation?}} we have:
\begin{equation}
  p(Z_D)\propto\mathcal{N}(\mathbb{E}[Z_0] - \mathbb{E}[Z_1], \mathbb{V}(Z_0) + \mathbb{V}(Z_1)).
\end{equation}
We then sample new observations until we are at least 95\% confident
that $Z_D\neq 0$. In other words, when $p(Z_D<0)<0.025$, we accept
$h=0$, and when $p(Z_D<0)>0.975$, we accept $h=1$.

@@

\subsection{Design}

@@

\section{Analysis}

@@

\subsubsection{Response times}

The average response time across all stimuli was \ExpTime{} seconds,
and the average response time for an individual stimulus ranged from
\ExpTimeMin{} to \ExpTimeMax{} seconds. The average accuracy across
all stimuli was \ExpAccuracy{}, though there were \ExpNumChance{}
individual stimuli (out of 720) for which people were not above
chance.

As expected, the minimum angle of rotation was significanty
monotonically correlated with average per-stimulus response times,
both for flipped pairs (\ExpThetaTimeCorrFlipped{}) and same pairs
(\ExpThetaTimeCorrSame{}). This was also the case for the number of
actions taken by the \Oc{} model (\OcThetaTimeCorr{}), the \Th{}
model (\ThThetaTimeCorr{}), and the \Bq{} model
(\BqThetaTimeCorr{}), but not the \Hc{} model
(\HcThetaTimeCorr{}). Fig.~\ref{fig:response-times}a

@@

\subsubsection{Accuracy}

Of the models, the \Oc{} and \Th{} models peformed perfectly,
with 100\% accuracy. The \Bq{} model was very accurate
overall (\BqAccuracy{}), though there were \BqNumChance{} stimuli for
which it was not above chance. The \Hc{} model was not
particularly accurate, and although it did perform above chance
overall (\HcAccuracy{}), there were \HcNumChance{} individual stimuli
for which it was not above chance.

@@

\subsubsection{Practice effects}

>>>>

\title{Modeling mental rotation}

@@

\TODO{fix title}

@@

One of our most astonishing cognitive feats is the ability to
envision, manipulate, and plan with objects -- all without actually
perceiving them. This capacity for ``mental simulation'' has been
widely studied and has sparked intense debate about the underlying
representation of mental images \cite<e.g.,>{Kosslyn:2009tj,
  Pylyshyn:2002vk}. Despite this vast literature, there have been
surprisingly few computational models of mental imagery.

@@

In the classic experiment by \citeA{Shepard1971}, participants viewed
images of three-dimensional objects and had to determine whether the
images depicted the same object (which differed by a rotation) or two
separate objects (which differed by a reflection and a rotation). The
result was that people's response times had a strong linear
correlation with the minimum angle of rotation. The conclusion was
that people were solving this task by ``mentally rotating'' the
objects until they were congruent.

However, there are some questions regarding this process of problem
solving that are not explained by the alignment hypothesis: how do
people know the axis around which to rotate the objects? If the axis
is known, how do people know which direction to rotate the objects?
And finally, how do people know when to stop rotating and give a
response?

Previous models of mental rotation have, for the most part, not
addressed these questions. \citeA{Kosslyn:1977tv} focus primarily on
simulating the visual simulation buffer rather than the decision of
\textit{what} should be simulated. Similarly, \citeA{Glasgow:1992tj}
are more concerned with the representation of imagery than the
process. Although \citeA{Anderson1978} emphasizes the importance of
considering both representation and process, he dismisses the problem
of determining the direction and axis of rotation as merely a
``technical difficulty''.

The only model (of which the authors are aware) that has more
seriously attempted to address these concerns is the parallel-process
model of mental rotation by \citeA{Funt:1983wn}. He determines the
direction and axis of rotation using the method-of-moments method;
once this target rotation computed, his model rotates one object until
the target rotation is reached, and then compares the two to see if
they are congruent.

More recently, evidence has come to light that people \textit{do not}
actually rotate until the objects are congruent. Eye-tracking studies
suggest people pay attention to certain salient features of the
object, rather than a holistic representation \cite{Just1976}, and 
\citeA{Yuille:1982tx} e.g. rotating away from near-perfect matches \cite{Gardony:2013gn}

@@

\subsection{A rational analysis of mental rotation}

@@

here we use simple 2d models as a starting point, but the general case
of solving this problem is very difficult, \TODO{cite shape matching
  literature}

@@

Formally, we denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is
generated by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$,
where $\theta$ is a rotation, $h=0$ is the hypothesis that the images
depict the same object, and $h=1$ is the hypothesis that the images
depict mirror-image objects. The posterior probability of each
hypothesis given the observed shapes is then:
\begin{equation}
  p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta, h)p(X_a)p(h)p(\theta)\dif\theta.
  \label{eq:posterior}
\end{equation}
Because we ultimately want to determine which hypothesis is more
likely, the quantity of interest is a ratio $\ell:=p(h=0\ \vert\ X_a,
X_b) / p(h=1\ \vert\ X_a, X_b)$ which (assuming the two hypotheses are
equally likely \textit{a priori}, and that all rotations are equally
likely) simplifies to:
\begin{equation}
  \ell = \frac{\int p(X_b\ \vert\ X_a, \theta, \hi)\dif\theta}{\int p(X_b\ \vert\ X_a, \theta, \hf)\dif\theta}.
  \label{eq:lh-ratio}
\end{equation}
If $\ell > 1$, then we accept the hypothesis that the images depict
the same object ($\hi$); if $\ell < 1$, then we accept the hypothesis
that the images depict flipped objects ($\hf$).

@@

The previous hypothesis in the literature is that people perform a
mental rotation in the shortest direction until a match is found, and
that determining this direction is an implementation detail \TODO{cite
  anderson?}. To reflect this hypothesis, we created an ``oracle''
model which knows the correct rotation \textit{a priori}, and simply
performs rotations in the shortest direction until it reaches the
point where the shapes are aligned.

@@

Unfortunately, it is not always clear what constitutes a good enough
``match''. In the current formulation of the problem, we know the
exact geometry of the shapes, and know that a linear transformation
exists which will perfectly align the shapes. However, more generally,
shapes could be more complex and/or three-dimensional, be subject to
perceptual uncertainty, or even be not exactly identical. In these
cases, the problem of choosing a threshold becomes much harder: it is
not clear how to choose a reasonable threshold \textit{a priori}, as
the global optimum will depend on many factors. Indeed, shape matching
is a non-trivial problem in computer vision, requiring sophisticated
algorithms to achieve reasonable performance
\cite<e.g.,>{Belongie:2002tj,Sebastian:2003vm}.

Thus, the \Th{} model will not be robust beyond this domain. One
obvious alternative would be to perform an exhaustive search or use a
global optimisation strategy. However, this would not result in the
linear response time found by \citeA{Shepard1971}. A second
alternative (which would potentially demonstrate the linear response
time result) could instead just perform a Hill Climbing (\Hc{}) search
for a \textit{local} maximum. Similar to the \Th{} model, we use the
(local) MLE estimate of $\theta$, as in Eq.~\ref{eq:mle-lh-ratio}.

@@

An alternative to the previous models is not to perform a search for
the MLE, but to approximate the integrals in Eq.~\ref{eq:lh-ratio} by
picking locations to sample which will most improve accuracy of the
estimate. \TODO{Really need to motivate this better: why do we need
  something so complex? Talk about how it's robust to the shape of the
  function, the scale, noise, etc. It's complicated, but it should be
  able to handle these things, so let's try it and see if it works at
  all. If it does, we can work towards somethign more succinct in the
  future.}

Rather than a point estimate of the ratio, or even a simple
Monte Carlo estimate, we can infer a distribution by placing priors
over the similarity functions themselves:
\begin{equation}
\ell = \frac{\int \left[\int S(\Xb, f(\Xa, \theta, \hi))p(\theta)\dif\theta\right] p(S)\dif S}{\int \left[\int S(\Xb, f(\Xa, \theta, \hf))p(\theta)\dif\theta\right] p(S)\dif S}
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use Gaussian Process (GP) priors, as in \citeA{Osborne:2012tm}
\TODO{probably need to describe better what a GP is}. Because our
domain is on a circle, we utilize a periodic kernel for the covariance
\cite{Rasmussen:2006vz}, i.e. $k(\theta,
\theta^\prime)=h^2\exp[-\frac{2}{w^2}\sin^2(\frac{1}{2}(\theta-\theta^\prime))]$.

@@

Revisiting our likelihood ratio, we now have:
\begin{equation}
p(\ell)=p\left(\frac{Z_0}{Z_1}\right)\approx\frac{\mathcal{N}(Z_0\ \vert\ \mathbb{E}[Z_0], \mathbb{V}(Z_0))}{\mathcal{N}(Z_1\ \vert\ \mathbb{E}[Z_1], \mathbb{V}(Z_1))}.
\end{equation}
This distribution does not have a nice form. However, we really are
just interested in whether $Z_0>Z_1$ or $Z_1>Z_0$: thus, rather than
computing $p(\ell)$, we use $Z_D=Z_0-Z_1$. Assuming $Z_0$ and $Z_1$
are independent,\footnote{This assumption is probably not
  correct. \TODO{explanation?}} we have:
\begin{equation}
  p(Z_D)\propto\mathcal{N}(\mathbb{E}[Z_0] - \mathbb{E}[Z_1], \mathbb{V}(Z_0) + \mathbb{V}(Z_1)).
\end{equation}
We then sample new observations until we are at least 95\% confident
that $Z_D\neq 0$. In other words, when $p(Z_D<0)<0.025$, we accept
$h=0$, and when $p(Z_D<0)>0.975$, we accept $h=1$.

@@

\subsection{Participants and Design}

@@

\subsection{General analysis}

@@

\subsection{Response times}

The average response time across all stimuli was \ExpTime{} seconds,
and the average response time for an individual stimulus ranged from
\ExpTimeMin{} to \ExpTimeMax{} seconds.  As expected, the minimum
angle of rotation was significanty monotonically correlated with
average per-stimulus response times, both for flipped pairs
(\ExpThetaTimeCorrFlipped{}) and same pairs
(\ExpThetaTimeCorrSame{}). This was also the case for the number of
actions taken by the \Oc{} model (\OcThetaTimeCorr{}), the \Th{} model
(\ThThetaTimeCorr{}), and the \Bq{} model (\BqThetaTimeCorr{}), but
not the \Hc{} model
(\HcThetaTimeCorr{}). Fig.~\ref{fig:response-times}a

@@

\subsection{Accuracy}

The average accuracy across all stimuli was \ExpAccuracy{}, though
there were \ExpNumChance{} individual stimuli (out of 720) for which
people were not above chance. Of the models, the \Oc{} and \Th{}
models peformed perfectly, with 100\% accuracy. The \Bq{} model was
very accurate overall (\BqAccuracy{}), though there were
\BqNumChance{} stimuli for which it was not above chance. The \Hc{}
model was not particularly accurate, and although it did perform above
chance overall (\HcAccuracy{}), there were \HcNumChance{} individual
stimuli for which it was not above chance.

@@

\subsection{Practice effects}
